Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-21 paper2.qmd
section:6; process:write

<!-- 3. The issues with Case V, and the BTL model -->
Although replacing the Normal with the Logistic distribution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], Thurstone acknowledged that the two remaining assumptions of Case V--and, by extension, of the BTL model--could pose significant problems when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927b]. In particular, he noted that while assuming equal discriminal dispersions and zero correlation for discriminal processes simplifies the trait measurement model, these assumptions may fail to reflect the complexity of CJ data [@Thurstone_1927a; @Andrich_1978; @Kelly_et_al_2022]. As a result, they can lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].

Furthermore, because Thurstone aimed primarily to produce a trait's "coarse scaling" and "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp. 269], his theory did not specify how to use the resulting trait estimates for statistical inference. The CJ tradition has attempted to address this gap by separating trait estimation from the hypothesis testing process. Specifically, CJ studies often rely on point estimates of the traits (i.e., the BTL scores)--or their transformations--to conduct statistical inference. Although separating these processes simplifies the analysis of CJ data, the statistical literature cautions against this practice, as it can introduce bias into the analysis and compromise the reliability of the hypothesis test [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

@@

<!-- 6.1 About the implications of the sampling and comparison mechanisms -->
The explicit integration of the sampling and comparison mechanisms within our conceptual model offers a novel perspective on the first research avenue. Specifically, it shows how these mechanisms influence the data-generating process by acting as missing data sources, determining which observations are part of the data sample. This new perspective invites the application of Little and Rubin's principled missing data framework [@Little_et_al_2020], enabling a more rigorous evaluation of existing claims about these mechanisms, their influence on CJ outcomes, and their implications for designing and evaluating more complex experimental setups.

@@

One prominent claim in the CJ literature is that Thurstone's model naturally handles non-random missing data without compromising the reliability or validity of trait estimates [@Bramley_2008]. However, it is easy to imagine scenarios where this assumption may not hold. For instance, selecting a particular group of judges could result in an incomplete representation of the trait's full range and complexity, as judgment accuracy varies with demographic factors [@Bartholomew_et_al_2020a; @Kelly_et_al_2022] and even between individuals [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020]. Consequently, judges' selection could affect the formation of a "shared consensus", potentially biasing trait estimates and distorting the observed trait distribution [@Deffner_et_al_2022]. It is therefore essential to determine whether such sampling biases occur--and if so, whether they undermine the validity of trait estimates and the robustness of statistical inferences. Our framework offers a rigorous setting to test the scenario.

@@

Moreover, when designing and evaluating more complex experimental setups, our framework highlights one key principle: the sampling and comparison mechanisms directly influence the CJ outcome. In this study, we deliberately structured these mechanisms to produce data that is *missing completely at random* (MCAR)--that is, the missingness is unrelated to any observed or unobserved variables, including the outcome [@Little_et_al_2020]. This design offers one key advantage: it generates simple random samples that satisfy the condition of ignorability, enabling researchers to legitimately ignore missing data during analysis without introducing bias [@Everitt_et_al_2010; @Kohler_et_al_2019].

However, many modern CJ applications involve more complex forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) mechanism [@Little_et_al_2020]--that arise from the chosen sampling and comparison methods. Two notable examples are adaptive comparative judgment designs [@Pollitt_2012a; @Pollitt_2012b] and case-control studies [@Everitt_et_al_2010]. In adaptive designs, an algorithm defines the comparison mechanism by dynamically selecting which stimulus pairs judges evaluate. These designs have produced mixed evidence: some studies report improved trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], while others suggest these gains may be artificially inflated [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022]. In contrast, case-control studies rely on sampling mechanisms linked to non-assignable variables that likely influence the trait of interest. For instance, @Boonen_et_al_2020 selected children based on hearing status, a non-assignable variable that likely affects speech quality [@Boonen_et_al_2023; @Rivera_et_al_2024]. Nevertheless, regardless of their missingness mechanism, both designs could benefit from explicitly articulating their underlying causal structures. Doing so would enable researchers to assess and test their impact on trait estimation and statistical inference. Our framework provides a rigorous foundation for such articulations.

@@

<!-- 6.2 About judges biases, judgment discrepancies, and the number of judges and judgments -->
Additionally, by explicitly incorporating judges' biases, our framework provides a novel approach to investigate the second research avenue. Specifically, it enables researchers to determine whether such biases occur and, if so, to assess their impact on the reliability and validity of trait estimation. Although some anecdotal evidence already suggests judges' biases exist [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a], further research is needed to address several key questions. For instance: Can CJ data be analyzed under the assumption of "sample-free" trait calibration--that is, the estimated trait difference for any two scripts does not depend on which other scripts they are compared with, implying an absence of judges' bias? [@Bramley_2008; @Andrich_1978] If this assumption does not hold, how many judges and judgments are required to ensure acceptable levels of trait reliability and validity? Do training and expertise influence judgment biases, and to what extent? [@Kelly_et_al_2022] Finally, what are the benefits of *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times? [@Lawson_2015]

>>>>

<!-- 3. The issues with Case V, and the BTL model -->
Although replacing the Normal with the Logistic distribution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], Thurstone acknowledged that the two remaining assumptions of Case V--and, by extension, of the BTL model--could pose significant problems when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927b]. Specifically, while assuming assuming equal discriminal dispersions and zero correlation for discriminal processes simplifies the trait measurement model, these assumptions may fail to reflect the complexity of CJ data [@Thurstone_1927a; @Andrich_1978; @Kelly_et_al_2022]. As a result, they can lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].

Furthermore, because Thurstone aimed primarily to produce a trait's "coarse scaling" and "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp. 269], his theory did not specify how to use the resulting trait estimates for statistical inference. The CJ tradition has attempted to address this gap by separating trait estimation from hypothesis testing. Specifically, CJ studies often rely on point estimates of the traits--typically the BTL scores or their transformations--to conduct statistical inference. Although this separation simplifies CJ data analysis, the statistical literature warns that this practice can introduce bias into the analysis and compromise the reliability of the hypothesis tests [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

@@

<!-- 6.1 About the implications of the sampling and comparison mechanisms -->
The explicit integration of the sampling and comparison mechanisms within our conceptual model offers a novel perspective on the first research avenue. Specifically, it shows how these mechanisms influence the data-generating process by acting as missing data sources, determining which observations are part of the data sample. This new perspective invites the application of Little and Rubin's [-@Little_et_al_2020] principled missing data framework, enabling a more rigorous evaluation of existing claims about these mechanisms, their influence on CJ outcomes, and their implications for designing and evaluating more complex experimental setups.

@@

One prominent claim in the CJ literature is that Thurstone's model naturally handles non-random missing data without compromising the reliability or validity of trait estimates [@Bramley_2008]. However, it is easy to imagine scenarios where this assumption may not hold. For instance, selecting a particular group of judges could result in an incomplete representation of the trait's full range and complexity, as judgment accuracy varies with demographic factors [@Bartholomew_et_al_2020a; @Kelly_et_al_2022] and even between individuals [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020]. Consequently, judges' selection could affect the formation of a "shared consensus", potentially biasing trait estimates and distorting the observed trait distribution [@Deffner_et_al_2022]. It is therefore essential to determine whether such sampling biases undermine the validity of trait estimates and the robustness of statistical inferences. Our framework offers a rigorous setting to test this scenario.

@@

Moreover, when designing and evaluating more complex experimental setups, our framework highlights one key principle: the sampling and comparison mechanisms directly influence the CJ outcome. In this study, we deliberately structured these mechanisms to produce data that is *missing completely at random* (MCAR)--that is, the missingness is unrelated to any observed or unobserved variables, including the outcome [@Little_et_al_2020]. This design offers one key advantage: it generates simple random samples that satisfy the condition of ignorability, enabling researchers to legitimately *ignore* missing data during analysis without introducing bias [@Everitt_et_al_2010; @Kohler_et_al_2019].

However, many modern CJ applications involve more complex forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) mechanism [@Little_et_al_2020]--that arise from the chosen sampling and comparison methods. Two notable examples are adaptive comparative judgment designs [@Pollitt_2012a; @Pollitt_2012b] and case-control studies [@Everitt_et_al_2010]. In adaptive designs, an algorithm defines the comparison mechanism by dynamically selecting which stimulus pairs judges evaluate. These designs have produced mixed evidence: some studies report improved trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], while others suggest these gains may be artificially inflated [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022]. In contrast, case-control studies rely on sampling mechanisms linked to non-assignable variables that likely influence the trait of interest. For instance, @Boonen_et_al_2020 selected children based on hearing status, a non-assignable variable that likely affects speech quality [@Boonen_et_al_2023; @Rivera_et_al_2024]. Nevertheless, regardless of their missingness mechanism, both designs could benefit from explicitly articulating their underlying causal structures. Doing so would enable researchers to assess and test their impact on trait estimation and statistical inference. Our framework, again, provides a rigorous foundation for such articulations.

@@

<!-- 6.2 About judges biases, judgment discrepancies, and the number of judges and judgments -->
Additionally, by explicitly incorporating judges' biases, our framework provides a novel approach to investigate the second research avenue. Specifically, it enables researchers to determine whether such biases occur and, if so, to assess their impact on the reliability and validity of trait estimation. Although some anecdotal evidence already suggests judges' biases exist [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a], further research is needed to address several critical questions. For instance, can CJ data be analyzed under the assumption of "sample-free" trait calibration, which implies an absence of judges' bias? [@Bramley_2008; @Andrich_1978] If this assumption does not hold, how many judges and judgments per judge are required to ensure acceptable levels of trait reliability and validity? Do training and expertise influence judgment biases, and to what extent? [@Kelly_et_al_2022] Finally, what are the benefits of *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times, in assessing judgment accuracy? [@Lawson_2015]
